Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
4.1

\paragraph{Promises}

The goal of our analysis is to discover argument types for functions,
builtins, and specials (R functions such as \{, \code{if}, etc.).  First,
recall that R wraps arguments in promises (Section~\ref{sec:R}).  Naively
intercepting function calls and recording argument types would result in
incomplete data: While we can dig into promises without forcing them, if the
innermost value is an unevaluated expression we are unable to probe it
without evaluating it, which may alter program behaviour.  To circumvent
this obstacle, we instead intercept the forcing of promises: when a promise
is forced, we identify the surrounding call, collect the type of the value,
and add it to the call trace associated with the current function call.
Once function execution completes, we collect information about the return
value, add it to the call trace, and save the completed call trace to be
written out later.

First, we wanted to ensure that \typetracer captured types for unevaluated
arguments.  Unfortunately, not every argument of a function is necessarily
used during every function execution, meaning that some arguments may remain
unevaluated (as they are never forced), and there are a few reasons for
this.

One possibility is that the argument was simply unused given the control
flow of this particular execution (e.g. if the argument is only needed in
one branch of a conditional expression).  This mainly occurs when arguments
have default values for some parameters that are not critical to every
execution of the function.  In these cases, we still want to collect the
type of the default value.

Another possibility is that a value is used {\it without being forced}, as
is the case when values are ``metaprogrammed''.  This is common practice in
e.g. the \code{tidyverse}~\footnote{\AT{either talk about it here or in
    background}} ecosystem, as they use the metaprogramming capabilities of
R to implement a DSL for data analysis.

To tackle possibly unevaluated arguments, we make an {\it initial guess} of
function argument types on function entry. Specifically, we collect types
for argument default values, and ascribe an ``unused'' type for unevaluated
expressions.  If an argument promise is later forced, we simply update the
recorded type for the argument.

\paragraph{Missing arguments}
Second, we wanted to account for formal arguments (arguments which are named
in the function definition) which received no values when the function was
called. This represents cases where a function was not called with all of
its arguments, and no default value was specified for missing arguments.  We
record a ``missing'' type for such argument.

There are two obvious ways to deal with missing arguments: type them as $\bot$
or as $\top$.  As we are performing a dynamic analysis based on package test
code, we conservatively type these arguments as $\top$, or \code{any}, as It is
impossible to know dynamically what all possible function behaviours are.
We discuss the prevalence of such any-typed arguments in
Section~\ref{sec:evaluation}.

\paragraph{Non-local returns}
Third, we had to contend with R's backend sometimes performing longjumps,
which discards current R calls from the call stack.  In order to ensure that
call traces are not lost when a longjump occurs, \typetracer intercepts the
unwinding process initiated by the longjump and mimics the functionality of
the functions being exited.  The R-dyntrace framework supplies the return
value during the unwinding process (if there is indeed such a value), so we
can capture it, analyze it, and include it in the call trace.  In certain
situations, however, no return value can be obtained (e.g. a longjump
occurred as part of some error handling), and then we simply record a
``jumped'' return value, which we can deal with in post-processing.

\paragraph{Implementation details.}
We primarily rely on eight callbacks: \k{closure\_entry}, \k{closure\_exit},
\k{builtin\_entry}, \k{builtin\_exit}, \k{special\_entry},\k{special\_exit},
\k{promise\_force\_entry}, and \k{promise\_force\_exit}.  The
function-related callbacks are used mainly for bookkeeping: the analysis is
notified that a construct has been entered by pushing the call onto a stack.
The calls themselves store a trace object, it is that object that holds the
type information. As R can perform single or multiple dispatch on function
arguments depending on their class, the relevant information is kept by the
{\tt \_entry} variants.

The primary functionality of \typetracer is in {\tt promise\_force\_exit}.
As arguments to functions in R are wrapped in promises, we delay our
reflection to promise force time.  In R, promises are essentially objects
with value and expression slots.

When a promise is created for a particular expression, that expression is
merely put into the expression slot of a promise object, and a special {\it
  unbound} value is loaded into the value slot (unless some default value is
specified, wherein it is loaded into the value slot---this is the case with
default argument values).  When the promise is forced, the expression is
evaluated in the forcing context, and the value is stored in the value slot
of the promise.  So to get the type of a function argument, we check the
value slot of the argument promise---recursively if the value itself is a
promise.

>>>>

\paragraph{Promises}
The fact that arguments are evaluated lazily, i.e. expression are packing
into promises and evaluated on first access, complicates the information
gathering.  For example, some promises remain un-evaluated, and it would be
erroneous to force them as they may cause side-effects that will affect the
programs execution.  To deal with unevaluated arguments, we make an initial
guess for each argument at fufnction entry.  If the promise is later forced,
we simply update the recorded type for the argument.

\paragraph{Missing arguments}
Parameters which receive no values when the function is called are termed
missing. This occurs when a function was called with too few arguments and
no default value was specified for missing arguments.  We record a
\k{missing} type for such argument.  There are two obvious ways to deal with
missing arguments: type them as \any or type them as some unit type.  As we
are performing a dynamic analysis, we conservatively type them as \any.

\paragraph{Non-local returns}
When a function exits with a longjump there is no return value to speak
of. In order to ensure that call traces are valid when a longjump occurs, we
intercept the unwinding process initiated by the longjump and mimics the
functionality of the functions being exited. When no return value is present
for a call, we record a special \k{jumped} return type.

\paragraph{Implementation details.}
We primarily rely on eight callbacks: \k{closure\_entry}, \k{closure\_exit},
\k{builtin\_entry}, \k{builtin\_exit}, \k{special\_entry},
\k{special\_exit}, \k{promise\_force\_entry}, and
\k{prom\-ise\_force\_exit}.  The function-related callbacks are used mainly
for bookkeeping: the analysis is notified that a construct has been entered
by pushing the call onto a stack.  The calls themselves store a trace
object, it is that object that holds the type information. As R can perform
single or multiple dispatch on function arguments depending on their class,
the relevant information is kept by the {\tt \_entry} variants.
